---
title: "Bias analysis"
output:
  html_document:
    toc: true
---

We investigate the simulations from the eXpress paper.

Let's just load some packages we need:

```{r}
library("data.table")
library("dplyr")
library("sleuth")
library("mamabear")
```

# Preliminaries

```{r}
base_dir <- "../../results"
```

kallisto results:

```{r}
kal_bb <- read_kallisto_h5(
  file.path(base_dir, "bias_and_error", "kallisto_bias", "abundance.h5"),
  FALSE)$abundance
kal_bn <- read_kallisto_h5(
  file.path(base_dir, "bias_and_error", "kallisto", "abundance.h5"),
  FALSE)$abundance
kal_eb <- read_kallisto_h5(
  file.path(base_dir, "error_only", "kallisto_bias", "abundance.h5"),
  FALSE)$abundance
kal_en <- read_kallisto_h5(
  file.path(base_dir, "error_only", "kallisto", "abundance.h5"),
  FALSE)$abundance
```

eXpress results were obtained from authors:

```{r}
xprs_bb <- read.table(
  gzfile(file.path(base_dir, "results_bias.xprs.gz")),
  stringsAsFactors = FALSE, header = TRUE) %>%
  mutate(tpm = (fpkm / sum(fpkm)) * 1e6)
# TODO: find out if express was run with bias correction on error only data
xprs_eb <- read.table(
  gzfile(file.path(base_dir, "results_error_only.xprs.gz")),
  stringsAsFactors = FALSE, header = TRUE) %>%
  mutate(tpm = (fpkm / sum(fpkm)) * 1e6)  
```

And now the oracles:

```{r}
reformat_oracle <- function(df, counts) {
  # assumes read a "transcripts.in" file
  colnames(df) <- c("target_id", "length", "rho")
  # the weird computation of tpm is to make it slightly more stable
  df <- df %>%
    mutate(rho = ifelse(rho < 0, 0, rho)) %>%
    mutate(
      tpm = (rho * 1e6 / sum(rho * 1e6)) * 1e6,
      rho = rho / sum(rho)
      )
  df <- left_join(df, counts, by = c("target_id"))
  df <- mutate(df, counts = ifelse(is.na(counts), 0, counts))
  
  df
}
```

```{r}
oracle_bias <- read.table("../../simdata/bias_and_error/transcripts.in", header = TRUE,
  stringsAsFactors = FALSE)
oracle_error <- read.table("../../simdata/error_only/transcripts.in", header = TRUE,
  stringsAsFactors = FALSE)
```

The true number of counts isn't provided, so instead we will derive this by reading the actual reads.

```{r}
oracle_counts_bias <- read.table("../../simdata/bias_and_error/counts.txt",
  header = FALSE, stringsAsFactors = FALSE)
colnames(oracle_counts_bias) <- c("target_id", "counts")
oracle_counts_error <- read.table("../../simdata/error_only/counts.txt",
  header = FALSE, stringsAsFactors = FALSE)
colnames(oracle_counts_error) <- c("target_id", "counts")
```

Let's now join the counts and reformat the oracle to be consistent with what `sleuth` is expecting:

```{r}
oracle_bias <- reformat_oracle(oracle_bias, oracle_counts_bias)
oracle_error <- reformat_oracle(oracle_error, oracle_counts_error)
```

## Merging results

We merge results using sleuth.

```{r}
mr_bias <- merge_results(
  list(kal_bb, kal_bn, xprs_bb),
  c("kallisto_bias", "kallisto", "express"),
  oracle_bias
  )
```

```{r}
mr_error <- merge_results(
  list(kal_eb, kal_en, xprs_eb),
  c("kallisto_bias", "kallisto", "express"),
  oracle_error
  )
```

# Bias and error results

```{r}
filtered_summary(mr_bias) %>%
  lapply(as.data.frame)
```

```{r}
filtered_summary(mr_bias, tpm_oracle > 1)  %>%
  lapply(as.data.frame)
```

# Error only results

```{r}
filtered_summary(mr_error) %>%
  lapply(as.data.frame)
```

```{r}
filtered_summary(mr_bias, tpm_oracle > 1)  %>%
  lapply(as.data.frame)
```

# Timings

```{r}
timing_fnames <- Sys.glob("../../benchmarks/*/*.json")
timings <- data.frame(
  kallisto_mode =  sub(".json", "", basename(timing_fnames)),
  dataset = basename(dirname(timing_fnames)),
  stringsAsFactors = FALSE)
timings$seconds <- sapply(timing_fnames, function(f) jsonlite::fromJSON(f)$wall_clock_time$s)
```

```{r}
timings <- timings %>%
  mutate(minutes = seconds / 60, hours = seconds / (60*60))
```

```{r}
timings %>% print(width = ncol(.))
```